---
title: "clustbench exploratory / fig 2 making"
author: "Izaskun Mallona"
output:
  html_document:
    theme: readable
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    highlight: tango
    keep_md: true
    self_contained: false
date: "`r format(Sys.Date(), '%B %d, %Y')`"
---


```{r setup, message = FALSE}
library(knitr)
library(tidyverse)
library(jsonlite)
library(ggplot2)
library(data.table)

knitr::opts_chunk$set(
  echo    = TRUE,
  warning = TRUE,
  message = TRUE,
  fig.width = 10,
  fig.height = 10,
  fig.path = "plots/",
  dev = c("png", "svg"),
  cache.lazy = FALSE,
  cache = TRUE)

```

```{r}

## given the way we have them, so we aim to provide 5 ks, middle being the true, but pad with 2s on the left
get_true_k <- function(x) {
  n <- length(x)
  if (n == 5) {
    return(x[3])
  } else if (n == 4) {
    return(x[3])
  } else if (n == 3) {
    return(x[2])
  } else {
    return(x[1])
  }
}


`%||%` <- function(a, b) if (!is.null(a)) a else b

flatten_parameters <- function(params, parameter_dir) {
  if (is.null(params) || length(params) == 0) {
    return(data.frame(parameter_dir = parameter_dir,
                      param_name    = NA_character_,
                      param_value   = NA_character_,
                      stringsAsFactors = FALSE))
  }
  flat <- unlist(params, use.names = TRUE)
  data.frame(
    parameter_dir = parameter_dir,
    param_name    = names(flat),
    param_value   = as.character(flat),
    stringsAsFactors = FALSE
  )
}

flatten_record <- function(rec) {
  mets <- rec$metrics
  perf <- rec$performance
  params_df <- flatten_parameters(rec$parameters, rec$parameter_dir)

  rows <- list()
  idx <- 0

  for (family in names(mets)) {
    fam_list <- mets[[family]]
    if (is.null(fam_list)) next

    for (metric_name in names(fam_list)) {
      metric_vals <- fam_list[[metric_name]]
      if (is.null(metric_vals)) next

      ks <- names(metric_vals)
      true_k <- get_true_k(ks)

      for (k in ks) {
        for (i in seq_len(nrow(params_df))) {
          idx <- idx + 1
          rows[[idx]] <- data.frame(
            backend       = rec$backend,
            seed          = rec$seed,
            run           = rec$run,
            generator     = rec$generator,
            dataset_name  = rec$dataset_name,
            method        = rec$method,          # keep "sklearn"
            method_full   = rec$method_full,     # "sklearn_method-birch" or "sklearn_method-kmeans"
            parameter_dir = rec$parameter_dir,   # "method-birch" or "method-kmeans"
            param_name    = params_df$param_name[i],
            param_value   = params_df$param_value[i],
            metric_family = family,
            metric_name   = metric_name,
            k             = as.integer(k),
            true_k        = as.integer(true_k),
            metric_value  = metric_vals[[k]],
            s             = perf$s,
            h_m_s         = perf[["h:m:s"]],
            max_rss       = perf$max_rss,
            max_vms       = perf$max_vms,
            max_uss       = perf$max_uss,
            max_pss       = perf$max_pss,
            io_in         = perf$io_in,
            io_out        = perf$io_out,
            mean_load     = perf$mean_load,
            cpu_time      = perf$cpu_time,
            stringsAsFactors = FALSE
          )
        }
      }
    }
  }

  do.call(rbind, rows)
}


```


```{r}

records <- fromJSON("aggregated_results.json", simplifyVector = FALSE)

fd_list <- vector("list", length(records))
for (i in seq_along(records)) {
  if (i == 1 || i == 2 || i == 100 || i %% 1000 == 0)
      cat("Processing record", i, "of", length(records), "\n")
  fd_list[[i]] <- flatten_record(records[[i]])
}
fd <- data.table::rbindlist(fd_list, use.names = TRUE, fill = TRUE)

## dim(fd)

## str(fd)
## table(is.na(fd$max_rss))
## table(fd$method_full)
## table(vapply(records, function(x) x$method_full, character(1)))

fd <- as.data.frame(fd)

fd$k <- as.integer(fd$k)
fd$true_k <- as.integer(fd$true_k)

cols_to_num <- c("max_rss","max_vms","max_uss","max_pss",
                 "io_in","io_out","mean_load","cpu_time")

fd[cols_to_num] <- lapply(fd[cols_to_num], function(x) {
  x[x == "NA"] <- NA_character_
  as.numeric(x)
})

fd$k_offset <- fd$k - fd$true_k
write.csv(fd, file = 'aggregated_results.csv')

```

# QC

Is the speed of computing the metric, or of running the method?

```{r}
print(ggplot(fd[fd$backend %in% c("conda","oras","envmodules"), ],
       aes(x = backend, y = cpu_time, fill = backend)) +
  geom_boxplot(outlier.alpha = 0.3) +
  theme_minimal(base_size = 14) +
  labs(title = "CPU time by backend",
       x = "Backend",
       y = "CPU Time (s)") +
  scale_fill_brewer(palette = "Set2"))
```

Clearly the method, so all good. With segments following dataset/method/params.

```{r}
print(ggplot(fd[fd$backend %in% c("conda","oras","envmodules"), ],
       aes(x = backend, y = cpu_time, fill = backend)) +
  geom_boxplot(outlier.alpha = 0.3) +
  # add points for each method/params combination
  geom_point(alpha = 0.6, position = position_jitter(width = 0.15)) +
  # connect points across backends for same module+params
  geom_line(aes(group = interaction(method, method_full, seed, run, generator, dataset_name, k)),
            alpha = 0.1, color = "grey40") +
  facet_wrap(~metric_name, scales = "free_y") +
  theme_minimal(base_size = 14) +
  labs(title = "CPU time by backend",
       x = "Backend",
       y = "CPU time (s)") +
  scale_color_brewer(palette = "Set1")
)
```

# Consistency checks

Are there cputimes that are redudant/incosistent for repeated runs?

```{r}
str(fd)

keys <- c("dataset_name", "method", "method_full", "generator",
          "backend", "k", "seed", "run")

fd_inconsistencies <- fd %>%
  group_by(across(all_of(keys))) %>%
  summarise(
    n_cpu_time = n_distinct(cpu_time, na.rm = TRUE),
    cpu_times  = list(unique(cpu_time)),
    .groups = "drop"
  ) %>%
  filter(n_cpu_time > 1)

stopifnot(nrow(fd_inconsistencies) == 0)
```

Seed consistencies

```{r}
keys <- c("dataset_name", "method", "method_full", "generator",
          "backend", "run", "k", "metric_name")

fd_seed_diff <- fd %>%
  group_by(across(all_of(keys))) %>%
  summarise(
    n_seeds = n_distinct(seed),
    n_metric_values = n_distinct(metric_value, na.rm = TRUE),
    metric_values = list(unique(metric_value)),
    .groups = "drop"
  ) %>%
  filter(n_seeds > 1 & n_metric_values > 1)

stopifnot(nrow(fd_seed_diff) == 0)
```

`k` inconsistencies


```{r}
stopifnot(range(fd$k_offset) == c(-2, 2))
```

Are seeds ok? why is `fcps_method-FCPS_MinEnergy_seed-2` a method name, which seed is that? **looks wrong**

```{r}
table(fd$method_full, fd$seed)
```




# CPU by backend

## Censoring aware : cpu_time < 0.05 plotted as 0.05

We impute NA cpu_time as 0.05 s

```{r}

min(fd$cpu_time, na.rm = TRUE)
fd$imputed_cpu_time <- ifelse(is.na(fd$cpu_time), no = fd$cpu_time, yes = 0.05)
fd$censored_cpu_time  <- is.na(fd$cpu_time)

## also, we average  for repeated runs, seeds and ks

fd_avg <- fd %>%
  filter(backend %in% c("conda","oras","envmodules")) %>%
  group_by(backend, method, method_full, generator, dataset_name) %>%
  summarise(imputed_cpu_time = mean(imputed_cpu_time, na.rm = TRUE),
            censored_cpu_time = any(censored_cpu_time),
            .groups = "drop")


## str(fd_avg)
## head(fd_avg)

ggplot(fd_avg, aes(x = backend, y = imputed_cpu_time)) +
  geom_boxplot(outlier.alpha = 0.3) +
  geom_point(aes(color = method_full, shape = censored_cpu_time),
             alpha = 0.6, position = position_jitter(width = 0.15)) +
  geom_line(aes(group = interaction(method, method_full, generator, dataset_name)),
            alpha = 0.2, color = "grey40") +
  theme_minimal(base_size = 14) +
  facet_wrap(~method, scales = "free_y") + 
  labs(title = "CPU time by backend - censored",
       x = "Backend",
       y = "CPU time (s)") +
  scale_y_sqrt()


ggplot(fd_avg, aes(x = backend, y = imputed_cpu_time)) +
  geom_boxplot(outlier.alpha = 0.3) +
  geom_point(aes(color = method_full, shape = censored_cpu_time),
             alpha = 0.6, position = position_jitter(width = 0.15)) +
  geom_line(aes(group = interaction(method, method_full, generator, dataset_name)),
            alpha = 0.2, color = "grey40") +
  theme_minimal(base_size = 14) +
  facet_wrap(~method, scales = "free_y") + 
  labs(title = "CPU time by backend - censored",
       x = "Backend",
       y = "CPU time (s)")

```

# Results consistency across backends

## Unreadable perf metrics plot

```{r, fig.width = 30, fig.height = 30}

print(
  ggplot(fd, aes(x = backend, y = metric_value)) +
  # boxplots still colored by backend
  geom_boxplot(aes(fill = backend), outlier.alpha = 0.3) +
  # points colored by run
  geom_point(aes(color = run),
             alpha = 0.6,
             position = position_jitter(width = 0.15)) +
  # connect points across backends for same method/params/run/etc.
  geom_line(aes(group = interaction(method, method_full, seed, run, generator, dataset_name, k),
                color = run),
            alpha = 0.2) +
  facet_wrap(dataset_name ~ metric_name, scales = "free_y") +
  theme_minimal(base_size = 14) +
  labs(title = "Metrics consistency across backends (stage = methods)",
       x = "Backend",
       y = "Metric Value") +
  scale_fill_brewer(palette = "Set1") +
  scale_color_brewer(palette = "Dark2"))

```

# Pairwise correlations

```{r}
df_run <- fd %>%
  select(dataset_name, method, k, metric_name, run, seed, backend, metric_value)


# a bit of collapsing here
wide_run <- fd %>%
  filter(backend %in% c("conda","oras","envmodules")) %>%
  group_by(dataset_name, method, k, metric_name, run, seed, backend) %>%
  summarise(metric_value = mean(as.numeric(metric_value), na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = backend, values_from = metric_value)

cors_run <- wide_run %>%
  group_by(metric_name) %>%
  summarise(
    cor_conda_oras   = cor(conda, oras, use="complete.obs"),
    cor_conda_envmodules  = cor(conda, envmodules, use="complete.obs"),
    cor_oras_envmodules = cor(oras, envmodules, use="complete.obs")
  )

print(cors_run)

```

# Computational performance metrics

## Non-complete obs

```{r, fig.width = 10, fig.height=10}

perf_metrics <- c("cpu_time","max_rss","max_vms","max_uss",
                  "max_pss","io_in","io_out","mean_load")

fd_long <- fd %>%
  filter(backend %in% c("conda","oras","envmodules")) %>%
  pivot_longer(cols = all_of(perf_metrics),
               names_to = "metric",
               values_to = "value")


#  boxplots + jittered scatter, faceted by metric
ggplot(fd_long, aes(x = backend, y = value, fill = backend)) +
  geom_boxplot(outlier.alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.1, size = 1, color = "black") +
  facet_wrap(~metric, scales = "free_y") +
  theme_minimal(base_size = 14) +
  labs(title = "Performance metrics by backend",
       x = "Backend",
       y = "Value") +
  scale_fill_brewer(palette = "Set2")
```

## Complete obs

Now the same but making sure only items that are measured across all backends (so no NAs because they're too quick and not profiled) are shown

```{r, fig.width = 10, fig.height=10}

fd_long_complete <- fd %>%
  filter(backend %in% c("conda","oras","envmodules")) %>%
  pivot_longer(cols = all_of(perf_metrics),
               names_to = "metric",
               values_to = "value") %>%
  group_by(method_full, dataset_name, seed, run, generator, k, metric) %>%
  filter(n_distinct(backend) == 3, !any(is.na(value))) %>%
  ungroup()

ggplot(fd_long_complete, aes(x = backend, y = value, fill = backend)) +
  geom_boxplot(outlier.alpha = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.1, size = 1, color = "black") +
  facet_wrap(~metric, scales = "free_y") +
  theme_minimal(base_size = 14) +
  labs(title = "Perf metrics by backend (only complete groups)",
       x = "backend",
       y = "perf value") +
  scale_fill_brewer(palette = "Set2")


```

## Non-complete observations colored by method provider, with a bug

Something wrong with method "1" etc here?


```{r, fig.width = 10, fig.height = 10}
fd_long <- fd %>%
  pivot_longer(cols = all_of(perf_metrics),
               names_to = "metric",
               values_to = "value")

## table(fd_long$method, fd_long$method_full)

ggplot(fd_long, aes(x = backend, y = value)) +
  # boxplots filled by backend
  ## geom_boxplot(aes(fill = backend), outlier.alpha = 0.3) +
  # points colored by method
  geom_point(aes(color = method),
             alpha = 0.6,
             position = position_jitter(width = 0.15)) +
  # connect points across backends for same method/run/etc.
  geom_line(aes(group = interaction(method, method_full, seed, run, generator, dataset_name, k),
                color = run),
            alpha = 0.1) +
  facet_wrap(~metric, scales = "free_y") +
  theme_minimal(base_size = 14) +
  labs(title = "Performance metrics by backend",
       x = "Backend",
       y = "Value") +
  scale_fill_brewer(palette = "Set2") +
  scale_color_brewer(palette = "Dark2")

```

# Choice of `k` - and its misspecification

## Deviation of performance at false k vs true k

Mind there is no such a thing as true k, we use the first labelset.

```{r}

fd_dev_k <- fd %>%
  group_by(method_full, dataset_name, metric_name) %>%
  # get the metric_value at the true_k for this group
  mutate(true_value = metric_value[k == true_k][1],
         deviation_k = metric_value - true_value) %>%
  ungroup()

summary(fd_dev_k$deviation_k)
## str(fd_dev_k)
ggplot(fd_dev_k, aes(x = factor(k), y = deviation_k, color = method_full)) +
  geom_point(alpha = 0.6, position = position_jitter(width = 0.15)) +
  facet_wrap(~ metric_name, scales = "free_y") +
  theme_minimal(base_size = 14) +
  labs(title = "Deviation of metric across ks",
       subtitle = "Deviation vs value at true k",
       x = "k",
       y = "Perf metric deviation from true k") +
  scale_color_brewer(palette = "Set1")
```

# Impact of seeds and repeated runs


What about the consistency across runs and seeds? e.g., do different seeds produce different results?

Only at true k.



```{r}

fd_true <- fd %>% filter(k == true_k)

fd_dev_seed <- fd_true %>%
  group_by(method_full, dataset_name, metric_name, run) %>%
  mutate(
    mean_true_val = mean(metric_value, na.rm = TRUE),
    deviation_seed = metric_value - mean_true_val
  ) %>%
  ungroup()

## floats strike again...
tol <- sqrt(.Machine$double.eps)
fd_dev_seed <- fd_dev_seed %>%
  mutate(deviation_seed = ifelse(abs(deviation_seed) < tol, 0, deviation_seed))


summary(fd_dev_seed)


ggplot(fd_dev_seed, aes(x = factor(seed), y = deviation_seed, color = method_full)) +
  geom_point(alpha = 0.6) +
  theme_minimal(base_size = 14) +
  labs(title = "Deviation at true k across seeds",
       x = "Seed",
       y = "Deviation from mean across seeds")


fd_dev_run <- fd_true %>%
  group_by(method_full, dataset_name, metric_name, seed) %>%
  mutate(
    mean_true_val = mean(metric_value, na.rm = TRUE),
    deviation_run = metric_value - mean_true_val
  ) %>%
  ungroup()

tol <- sqrt(.Machine$double.eps)
fd_dev_run <- fd_dev_run %>%
  mutate(deviation_run = ifelse(abs(deviation_run) < tol, 0, deviation_run))

summary(fd_dev_run)

```

# Comp vs clustering performance trade-offs


```{r, fig.width = 12}
## print(ggplot(fd_long, aes(x = backend, y = value, fill = backend)) +
##   geom_boxplot(outlier.alpha = 0.3) +
##   geom_jitter(width = 0.2, alpha = 0.1, size = 1) +
##   facet_wrap(~metric, scales = "free_y") +
##   theme_minimal(base_size = 14) +
##   labs(title = "comp perf by backend") +
##   scale_fill_brewer(palette = "Set2"))

## print(ggplot(fd_seed_summary,
##              aes(x = seed, y = mean_value,
##                  group = interaction(method_full, dataset_name, k),
##                  color = method_full)) +
##   geom_line(alpha = 0.5) +
##   geom_point(size = 2) +
##   geom_errorbar(aes(ymin = mean_value - sd_runs,
##                     ymax = mean_value + sd_runs),
##                 width = 0.2, alpha = 0.4) +
##   facet_wrap(~metric_name, scales = "free_y") +
##   theme_minimal(base_size = 14) +
##   labs(title = "Seed sensitivity",
##        subtitle = "Error bars depict runs variability"))

print(ggplot(fd, aes(x = cpu_time, y = metric_value,
                     color = method_full, shape = backend)) +
  geom_point(alpha = 0.6) +
  facet_wrap(~metric_name, scales = "free_y") +
  theme_minimal(base_size = 14) +
  labs(title = "clustering metrics vs runtime trade‑offs",
       x = "CPU time (s)",
       y = "clustering metric value"))

print(ggplot(fd, aes(x = max_rss, y = metric_value,
                     color = method_full, shape = backend)) +
  geom_point(alpha = 0.6) +
  facet_wrap(~metric_name, scales = "free_y") +
  theme_minimal(base_size = 14) +
  labs(title = "clustering metrics vs RSS trade‑offs",
       x = "max RSS (MB)",
       y = "clustering metric value"))

```

# Bland Altmans for ks vs true k

Again there is no such a thing as a true k

## Colored by method

```{r, fig.width = 15, fig.height = 15}
fd_dev_true <- fd %>%
  group_by(method_full, dataset_name, metric_name) %>%
  # get the metric value at true_k
  mutate(true_value = metric_value[k == true_k][1],
         diff_val   = metric_value - true_value,
         mean_val   = (metric_value + true_value)/2) %>%
  ungroup()

# Bland–Altman style plot: any k vs true_k
ggplot(fd_dev_true, aes(x = mean_val, y = diff_val, color = method_full)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = mean(fd_dev_true$diff_val, na.rm = TRUE),
             linetype = "dashed", color = "blue") +
  geom_hline(yintercept = mean(fd_dev_true$diff_val, na.rm = TRUE) + 1.96*sd(fd_dev_true$diff_val, na.rm = TRUE),
             linetype = "dotted", color = "red") +
  geom_hline(yintercept = mean(fd_dev_true$diff_val, na.rm = TRUE) - 1.96*sd(fd_dev_true$diff_val, na.rm = TRUE),
             linetype = "dotted", color = "red") +
  facet_wrap(~metric_name, scales = "free") +
  theme_minimal(base_size = 14) +
  labs(title = "Bland–Altman: any k vs true_k",
       x = "Mean of k and true_k",
       y = "Difference (k - true_k)")
```

## Colored by dataset

```{r, fig.width = 15, fig.height = 15}
ggplot(fd_dev_true, aes(x = mean_val, y = diff_val, color = dataset_name)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = mean(fd_dev_true$diff_val, na.rm = TRUE),
             linetype = "dashed", color = "blue") +
  geom_hline(yintercept = mean(fd_dev_true$diff_val, na.rm = TRUE) + 1.96*sd(fd_dev_true$diff_val, na.rm = TRUE),
             linetype = "dotted", color = "red") +
  geom_hline(yintercept = mean(fd_dev_true$diff_val, na.rm = TRUE) - 1.96*sd(fd_dev_true$diff_val, na.rm = TRUE),
             linetype = "dotted", color = "red") +
  facet_wrap(~metric_name, scales = "free") +
  theme_minimal(base_size = 14) +
  labs(title = "Bland–Altman: any k vs true_k",
       x = "Mean of k and true_k",
       y = "Difference (k - true_k)")



```

## Grid of Bland Altmans - misspecified k

More Bland Altmans

```{r, fig.width = 50, fig.height = 30}
ggplot(fd_dev_true,
       aes(x = mean_val, y = diff_val, color = method_full)) +   # color by method_full
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = mean(fd_dev_true$diff_val, na.rm = TRUE),
             linetype = "dashed", color = "blue") +
  geom_hline(yintercept = mean(fd_dev_true$diff_val, na.rm = TRUE) + 
                           1.96*sd(fd_dev_true$diff_val, na.rm = TRUE),
             linetype = "dotted", color = "red") +
  geom_hline(yintercept = mean(fd_dev_true$diff_val, na.rm = TRUE) - 
                           1.96*sd(fd_dev_true$diff_val, na.rm = TRUE),
             linetype = "dotted", color = "red") +
  facet_grid(metric_name ~ dataset_name, scales = "free") +
  theme_minimal(base_size = 14) +
  labs(title = "Bland–Altman: any k vs true_k",
       x = "Mean of k and true_k",
       y = "Difference (k - true_k)")

```

## Effects of misspecifying `k`, unreadable QC plot

```{r, fig.width = 30, fig.height = 30}
true_vals <- fd %>%
  filter(k == true_k) %>%
  select(method_full, dataset_name, metric_name, run, seed, true_value = metric_value)

fd_dev_k <- fd %>%
  group_by(method_full, dataset_name, metric_name, run, seed) %>%
  mutate(true_value = first(metric_value[k == true_k]),
         deviation_k = metric_value - true_value) %>%
  ungroup()

ggplot(fd_dev_k, aes(x = factor(k), y = deviation_k, color = method_full)) +
  geom_point(alpha = 0.6, position = position_jitter(width = 0.15)) +
  facet_grid(dataset_name ~ metric_name, scales = "free_y") +
  theme_minimal(base_size = 14) +
  labs(title = "Deviation of metric across ks",
       subtitle = "Deviation vs value at true k",
       x = "k",
       y = "Perf metric deviation from true k") +
  scale_color_brewer(palette = "Set1")

```
